Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-05-03 writeups/Su_weak_tides.tex
stash 05/03 bit more before meeting to do

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
        xcolor,
    }

@@

We consider a star of mass $M_\star$ hosting an inner oblate planet of mass $m$
and radius $R$ on a circular orbit with semi-major axis $a$ and an outer
perturber of mass $m_{\rm p}$ on a circular orbit with semi-major axis $a_{\rm
p}$. We assume that the two orbits are mildly misaligned with mutual inclination
$I$. Denote $\bm{S}$ the spin angular momentum and $\bm{L}$ the orbital angular
momentum of the planet, and $\bm{L}_{\rm p}$ the angular momentum of the
perturber. The corresponding unit vectors are $\uv{s} \equiv \bm{S} / S$,
$\uv{l} \equiv \bm{L} / L$, and $\uv{l}_{\rm p} \equiv \bm{L}_{\rm p} / L_{\rm
p}$. The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disk axis $\uv{l}_{\rm p}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm p}$, so $\uv{l}_{\rm
p}$ and $\uv{l}$ are nearly constant. The equations of motion for $\uv{s}$ and
$\uv{l}$ in this limit are \citep{anderson2018teeter, su2020}
\begin{align}
    \rd{\uv{s}}{t}
        &= \omega_{\rm sl}\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times \uv{l}}
        \equiv \alpha\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times
        \uv{l}},\label{eq:dsdt1}\\
    \rd{\uv{l}}{t} &= \omega_{\rm lp}\p{\uv{l} \cdot \uv{l}_{\rm p}}\p{\uv{l}
        \times \uv{l}_{\rm p}} \equiv -g\p{\uv{l} \times \uv{l}_{\rm p}},
        \label{eq:dldt1}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 mR^2 M_\star}{2a^3 I\Omega_{\rm s}}
        = \frac{3k_q}{2k}\frac{M_\star}{m}\p{\frac{R}{a}}^3 \Omega_{\rm s},
            \label{eq:wsl}\\
    \omega_{\rm lp} &= \frac{3m_p}{4M_\star}\p{\frac{a}{a_{\rm p}}}^3
        n.\label{eq:wlp}
\end{align}
In Eq.~\eqref{eq:wsl}, $\Omega_{\rm s}$ is the spin frequency of the inner
planet, $I = k mR^2$ (with $k$ a constant) is its moment of inertia and $J_2 =
k_{\rm q}\Omega_{\rm s}^2 (R^3/Gm)$ (with $k_{q}$ a constant) is its rotation-induced
(dimensionless) quadrupole moment [for a body with uniform density, $k=0.4,
k_{\rm q}=0.5$; for rocky planets, $k\simeq 0.2$ and $k_{\rm q}\simeq 0.17$
\citep[e.g.][]{lainey2016quantification} \textcolor{red}{? not sure}]. In other
studies, $3k_{\rm q} / 2 k$ is often notated as $k_2 / 2C$
\citep[e.g.][]{millholland_disk}. In Eq.~\eqref{eq:wlp}, $n \equiv
\sqrt{GM_\star/a^3}$ is the inner planet's orbital mean motion,  and we have
assumed $a_{\rm p}\gg a$ and included only the leading-order (quadrupole)
interaction between the inner planet and perturber. Following standard notation,
we define $\alpha = \omega_{\rm sl}$ and $g \equiv -\omega_{\rm 1p} \cos I$
\citep[e.g.][]{colombo1966}.

@@

The equilibria of Eq.~\eqref{eq:dsdt_rot} are referred to as \emph{Cassini
States} \citep[CSs;][]{colombo1966, peale1969}. We follow the notation of
\citet{su2020}, where the parameter
\begin{equation}
    \eta \equiv -\frac{g}{\alpha},\label{eq:def_eta}
\end{equation}
is used. For a given value of $\eta$, there can be either two or four CSs, all
of which require $\uv{s}$ lie in the plane of $\uv{l}$ and $\uv{l}_{\rm p}$.
Following the standard convention and nomenclature, CSs 1, 3, and 4 have $\phi =
0$ and $\theta < 0$, while CS2 has $\phi = \pi$ and $\theta > 0$.
Figure~\ref{fig:cs_locs} shows the CS obliquities as a function of $\eta$, each
of which satisfies
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.\label{eq:cs_zero}
\end{equation}
Note that there are four CSs when $\eta < \eta_{\rm c}$ and only two when $\eta
> \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2}.\label{eq:def_etac}
\end{equation}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini State obliquities $\theta$ as a function of $\eta \equiv
    -g/\alpha$ (Eq.~\ref{eq:def_eta}) for $I = 5^\circ$. The vertical dashed
    line denotes $\eta_{\rm c}$, where the number of Cassini States changes from
    four to just two (Eq.~\ref{eq:def_etac}). The horizontal dashed line gives
    $\theta = I$, the asymptotic value of Cassini State 2's obliquity when $\eta
    \to \infty$.}\label{fig:cs_locs}
\end{figure}

The Hamiltonian corresponding to Eq.~\eqref{eq:dsdt_rot} is
\begin{align}
    H &= -\frac{\alpha}{2}\p{\uv{s} \cdot \uv{l}}^2
            - g\p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{\alpha}{2} \cos^2\theta
            - g\p{\cos\theta \cos I - \sin I \sin\theta \cos \phi}.\label{eq:H}
\end{align}
Here, $\cos \theta$ and $\phi$ form a canonically conjugate pair of variables.
Figure~\ref{fig:1contours} shows the level curves of this Hamiltonian for $I =
20^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:def_etac}). When $\eta
< \eta_{\rm c}$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in the
phase space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. Trajectories in zone
II librate about CS2 while those in zones I and III circulate. On the other
hand, when $\eta > \eta_{\rm c}$, all trajectories circulate. When the
separatrix exists, we divide it into two curves: $\mathcal{C}_+$, the boundary
between zones I and II, and $\mathcal{C}_-$, the boundary between zones II and
III\@.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/1contours.png}
    \caption{Level curves of the Cassini State Hamiltonian (Eq.~\ref{eq:H}) for
    $I = 5^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:def_etac}).
    For $\eta < \eta_{\rm c}$, there are four Cassini States (labeled), while
    for $\eta > \eta_{\rm c}$ there are only two. In the former case, the
    existence of a \emph{separatrix} (solid black lines) separates phase space
    into three numbered zones (I/II/III, labeled). Finally, we denote the upper
    and lower legs of the separatrix by $\mathcal{C}_{\pm}$ respectively, as
    shown in the upper two panels.}\label{fig:1contours}
\end{figure}

@@

If the alignment torque is weak ($\abs{gt_{\rm s}} \gg 1$), then the fixed points of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) are just slightly shifted CSs. This
shift can be calculated cleanly to leading order: all of the CS obliquities
$\theta_{\rm cs}$ are unchanged while the azimuthal angle $\phi_{\rm cs}$ for
each CS now satisfies
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin\theta_{\rm cs}}{\sin I \abs{g}t_{\rm s}}.
        \label{eq:mcs_shift}
\end{equation}
We can see that if $t_{\rm s} > t_{\rm s, c}$, where
\begin{align}
    t_{\rm s, c} &= \frac{1}{\abs{g}\sin I},
\end{align}
then Eq.~\eqref{eq:mcs_shift} will always have solutions for $\phi_{\rm cs}$,
and the alignment torque never changes the number of fixed points of the system.
If $t_{\rm s}$ is decreased below $t_{\rm s, c}$, CS2 and CS4 cease to be
fixed points if $\eta \ll 1$ \citep[as first noted in][]{fabrycky_otides}, as
$\theta_{\rm cs} \approx 90^\circ$ for these (see Fig.~\ref{fig:cs_locs}), while
the other CSs have small $\sin \theta_{\rm cs}$ and are only slightly perturbed.
Figure~\ref{fig:mcs} shows the obliquity and azimuthal angles for each of the
CSs in the $\eta \ll 1$ case, where it can be seen that CS2 and CS4 collide and
annihilate.. For the remainder of this section, we will consider the case where
$t_{\rm s} \gg t_{\rm s, c}$ and the CSs only differ slightly from their
unperturbed locations.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/0_stab.png}
    \caption{Modified CSs for $I = 5^\circ$ and $\eta = 0.2$, and the CS1 and
    CS3 obliquities have been offset (labeled in legend) to improve clarity of
    the plot.}\label{fig:mcs}
\end{figure}

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak tidal alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm s}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{align}
    0 &= \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm s}}}\lambda - \lambda_0^2,\\
    \lambda_0^2 &\equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}},\label{eq:def_l0_sq}\\
    \lambda &\approx -\frac{\cos \theta_{\rm cs}}{t_{\rm s}} \pm \sqrt{\lambda_0^2}.
        \label{eq:lambda_approx}
\end{align}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in, \citet{su2020} and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. From Figs.~\ref{fig:cs_locs}, we thus see that CSs 1 and
2 are stable while CS3 is unstable under the alignment torque. These
calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta
    \equiv \abs{g} / \alpha$ for the four CSs. The solid lines give
    $\lambda_0^2$ for the unperturbed $\phi_{\rm cs}$, and the dashed lines give
    the values for $\phi_{\rm cs}$ shifted by $60^\circ$ ($\phi_{\rm cs} =
    120^\circ$ for CS2 and $\phi_{\rm cs} = 60^\circ$ for CSs 1, 3, and 4).
    }\label{fig:lambda_full}
\end{figure}

@@

With the above results, we are equipped to ask questions about the dynamics of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}): what is the long term behavior for
a general initial $\uv{s}$? If $\eta > \eta_{\rm c}$, then the only possible
final spin state is CS2, and all initial conditions will evolve asympotically
towards it. As such, we consider only $\eta < \eta_{\rm c}$, where an initial
condition can asymptotically evolve towards either CS1 or CS2. In
Fig.~\ref{fig:toy_phop}, we show which CS an initial $\theta_0$ and $\phi_0$
evolves into, where we have taken $\abs{g}t_{\rm s} = 10^{-3}$, $\eta = 0.2$,
and $I = 20^\circ$. It is clear that initial $\uv{s}$ in zone I (following the
naming in Fig.~\ref{fig:1contours}) evolve into CS1, those in zone II evolve
into CS2, while those in zone III have a probabilistic outcome. We aim to
understand each of these in turn:
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{Plot illustrating the asymptotic behavior of initial conditions for
    $\eta = 0.2$ and $I = 20^\circ$. Each dot represents an initial spin
    orientation, and the coloring of the dot indicates which Cassini State
    (legend) the system asymptotes towards.}\label{fig:toy_phop}
\end{figure}

@@

To understand this process quantitatively, as well as associate probabilities to
the two possible outcomes, it is important to be more quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves with the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx -\alpha\sin I - g\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we can compute $\Delta H_{\pm}$ the
change over both legs of the separatrix with
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm s}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
Thus, if we evaluate $H$ every time that a trajectory crosses $\phi = 0$, we see
that it will initially be negative and increase for each circulation cycle until
the system encounters the separatrix where $H - H_{\rm sep} < \Delta H_-$.

>>>>

% vim 0_eta/1*.py 0_eta/3*.py 0_eta/6*.py 3_toy3/1sim.py 99_misc/0*.py 99_misc/2*.py 99_misc/5*.py 1_weaktide/5*.py 1_weaktide/6*.py
% cd 0_eta && python 1*.py && python 3*.py && python 6*.py && cd .. && cd 3_toy3 && python 1sim.py && cd .. && cd 99_misc && python 0*.py && python 2*.py && python 5*.py && cd .. && cd 1_weaktide && python 5*.py && python 6*.py
    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
        xcolor,
    }

@@

We consider a star of mass $M_\star$ hosting an inner oblate planet of mass $m$
and radius $R$ on a circular orbit with semi-major axis $a$ and an outer
perturber of mass $m_{\rm p}$ on a circular orbit with semi-major axis $a_{\rm
p}$. We assume that the two orbits are mildly misaligned with mutual inclination
$I$. Denote $\bm{S}$ the spin angular momentum and $\bm{L}$ the orbital angular
momentum of the planet, and $\bm{L}_{\rm p}$ the angular momentum of the
perturber. The corresponding unit vectors are $\uv{s} \equiv \bm{S} / S$,
$\uv{l} \equiv \bm{L} / L$, and $\uv{l}_{\rm p} \equiv \bm{L}_{\rm p} / L_{\rm
p}$. The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disk axis $\uv{l}_{\rm p}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm p}$, so $\uv{l}_{\rm
p}$ and $\uv{l}$ are nearly constant. The equations of motion for $\uv{s}$ and
$\uv{l}$ in this limit are \citep{anderson2018teeter, su2020}
\begin{align}
    \rd{\uv{s}}{t}
        &= \omega_{\rm sl}\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times \uv{l}}
        \equiv \alpha\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times
        \uv{l}},\label{eq:dsdt1}\\
    \rd{\uv{l}}{t} &= \omega_{\rm lp}\p{\uv{l} \cdot \uv{l}_{\rm p}}\p{\uv{l}
        \times \uv{l}_{\rm p}} \equiv -g\p{\uv{l} \times \uv{l}_{\rm p}},
        \label{eq:dldt1}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 mR^2 M_\star}{2a^3 I\Omega_{\rm s}}
        = \frac{3k_q}{2k}\frac{M_\star}{m}\p{\frac{R}{a}}^3 \Omega_{\rm s},
            \label{eq:wsl}\\
    \omega_{\rm lp} &= \frac{3m_{\rm p}}{4M_\star}\p{\frac{a}{a_{\rm p}}}^3
        n.\label{eq:wlp}
\end{align}
In Eq.~\eqref{eq:wsl}, $\Omega_{\rm s}$ is the spin frequency of the inner
planet, $I = k mR^2$ (with $k$ a constant) is its moment of inertia and $J_2 =
k_{\rm q}\Omega_{\rm s}^2 (R^3/Gm)$ (with $k_{q}$ a constant) is its rotation-induced
(dimensionless) quadrupole moment [for a body with uniform density, $k=0.4,
k_{\rm q}=0.5$; for rocky planets, $k\simeq 0.2$ and $k_{\rm q}\simeq 0.17$
\citep[e.g.][]{lainey2016quantification} \textcolor{red}{? not sure}]. In other
studies, $3k_{\rm q} / 2 k$ is often notated as $k_2 / 2C$
\citep[e.g.][]{millholland_disk}. In Eq.~\eqref{eq:wlp}, $n \equiv
\sqrt{GM_\star/a^3}$ is the inner planet's orbital mean motion,  and we have
assumed $a_{\rm p}\gg a$ and included only the leading-order (quadrupole)
interaction between the inner planet and perturber. Following standard notation,
we define $\alpha = \omega_{\rm sl}$ and $g \equiv -\omega_{\rm 1p} \cos I$
\citep[e.g.][]{colombo1966}.

@@

The equilibria of Eq.~\eqref{eq:dsdt_rot} are referred to as \emph{Cassini
States} \citep[CSs;][]{colombo1966, peale1969}. We follow the notation of
\citet{su2020}, where the parameter
\begin{equation}
    \eta \equiv -\frac{g}{\alpha},\label{eq:def_eta}
\end{equation}
is used. For a given value of $\eta$, there can be either two or four CSs, all
of which require $\uv{s}$ lie in the plane of $\uv{l}$ and $\uv{l}_{\rm p}$.
Following the standard nomenclature, $\uv{s}$ and
$\uv{l}_{\rm p}$ are on opposite sides of $\uv{l}$ for CSs 1, 3, and 4, and are
on the same side for CS2. We depart from the standard convention
and simply label the CSs in spherical coordinates: figure~\ref{fig:cs_locs}
shows the CS obliquities as a function of $\eta$. CS1 and CS4 do not exist when
$\eta > \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2}.
        \label{eq:def_etac}
\end{equation}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs_phi.png}
    \caption{Cassini State obliquities $\theta$ as a function of $\eta \equiv
    -g/\alpha$ (Eq.~\ref{eq:def_eta}) for $I = 20^\circ$. The vertical dashed
    line denotes $\eta_{\rm c}$, where the number of Cassini States changes from
    four to just two (Eq.~\ref{eq:def_etac}). The y-axis labels on the right of
    the plot show the asymptotic obliquities for CS2 and CS3, $I$ and $180^\circ
    - I$ respectively. Note that $\theta$ does not follow the standard
    convention \citep[e.g.][]{colombo1966} and is simply the angle between
    $\uv{s}$ and $\uv{l}$.}\label{fig:cs_locs}
\end{figure}

The Hamiltonian corresponding to Eq.~\eqref{eq:dsdt_rot} is
\begin{align}
    H &= -\frac{\alpha}{2}\p{\uv{s} \cdot \uv{l}}^2
            - g\p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{\alpha}{2} \cos^2\theta
            - g\p{\cos\theta \cos I - \sin I \sin\theta \cos \phi}.\label{eq:H}
\end{align}
Here, $\cos \theta$ and $\phi$ form a canonically conjugate pair of variables.
Figure~\ref{fig:1contours} shows the level curves of this Hamiltonian for $I =
20^\circ$, for which $\eta_{\rm c} \approx 0.574$ (Eq.~\ref{eq:def_etac}). When $\eta
< \eta_{\rm c}$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in the
phase space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. Trajectories in zone
II librate about CS2 while those in zones I and III circulate. On the other
hand, when $\eta > \eta_{\rm c}$, all trajectories circulate. When the
separatrix exists, we divide it into two curves: $\mathcal{C}_+$, the boundary
between zones I and II, and $\mathcal{C}_-$, the boundary between zones II and
III\@.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/1contours20.png}
    \caption{Level curves of the Cassini State Hamiltonian (Eq.~\ref{eq:H}) for
    $I = 20^\circ$, for which $\eta_{\rm c} \approx 0.57$
    (Eq.~\ref{eq:def_etac}). For $\eta < \eta_{\rm c}$, there are four Cassini
    States (labeled), while for $\eta > \eta_{\rm c}$ there are only two. In the
    former case, the existence of a \emph{separatrix} (solid black lines)
    separates phase space into three numbered zones (I/II/III, labeled). We
    denote the upper and lower legs of the separatrix by $\mathcal{C}_{\pm}$
    respectively, as shown in the upper two panels. }\label{fig:1contours}
\end{figure}

@@

If the alignment torque is weak ($\abs{gt_{\rm s}} \gg 1$), then the fixed points of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) are just slightly shifted CSs. This
shift can be calculated cleanly to leading order: all of the CS obliquities
$\theta_{\rm cs}$ are unchanged while the azimuthal angle $\phi_{\rm cs}$ for
each CS now satisfies
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin\theta_{\rm cs}}{\sin I \abs{g}t_{\rm s}}.
        \label{eq:mcs_shift}
\end{equation}
We can see that if $t_{\rm s} > t_{\rm s, c}$, where
\begin{align}
    t_{\rm s, c} &= \frac{1}{\abs{g}\sin I},\label{eq:mcs_shift_crit}
\end{align}
then Eq.~\eqref{eq:mcs_shift} will always have solutions for $\phi_{\rm cs}$,
and the alignment torque never changes the number of fixed points of the system.
If $t_{\rm s}$ is decreased below $t_{\rm s, c}$, CS2 and CS4 cease to be
fixed points if $\eta \ll 1$ \citep[as first noted in][]{fabrycky_otides}, as
$\theta_{\rm cs} \approx 90^\circ$ for these (see Fig.~\ref{fig:cs_locs}), while
the other CSs have small $\sin \theta_{\rm cs}$ and are only slightly perturbed.
Figure~\ref{fig:mcs} shows the obliquity and azimuthal angles for each of the
CSs in the $\eta \ll 1$ case, where it can be seen that CS2 and CS4 collide and
annihilate. The phase shift $\phi_{\rm cs}$ for CS2 and CS4 for $t_{\rm s} >
t_{\rm s, c}$ can be predicted to good accuracy using Eq.~\eqref{eq:mcs_shift}
using $\theta_{\rm cs} \approx \pi/2 - \eta \cos (I) \approx 79^\circ$
\citep{su2020} for both, shown as the dashed lines in the bottom panel of
Fig.~\ref{fig:mcs}. For the remainder of this section, we will consider the case
where $t_{\rm s} \gg t_{\rm s, c}$ and the CSs only differ slightly from their
unperturbed locations.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/0_stab.png}
    \caption{Modified CS obliquities (top) and azimuthal angles (bottom) for $I
    = 20^\circ$ and $\eta = 0.2$, where the CS1 and CS3 obliquities have been
    offset (labeled in top legend) to improve clarity of the plot. In both
    panels, the solid lines give the result when applying a numerical root
    finding algorithm to the full equations of motion,
    Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}), while the dotted lines in the
    bottom panel give the CS2 and CS4 azimuthal angles according to
    Eq.~\eqref{eq:mcs_shift}. At $\abs{gt_{\rm s} \sin I} = 1$, CS2 and CS4
    collide and annihilate (Eq.~\ref{eq:mcs_shift_crit}).}\label{fig:mcs}
\end{figure}

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak tidal alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm s}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{equation}
    0 = \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm s}}}\lambda
        - \lambda_0^2,\label{eq:lambda_orig}
\end{equation}
where
\begin{equation}
    \lambda_0^2 \equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}}\label{eq:def_l0_sq}.
\end{equation}
When $t_{\rm s}$ is large, we can simplify Eq.~\eqref{eq:lambda_orig} to
\begin{equation}
    \lambda \approx -\frac{\cos \theta_{\rm cs}}{t_{\rm s}}
        \pm \sqrt{\lambda_0^2}. \label{eq:lambda_approx}
\end{equation}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in, \citet{su2020} and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. Evaluating at each of the CSs (see
Fig.~\ref{fig:cs_locs}), we conclude that CS1 and CS2 are stable and attracting
while CS3, stable in the absence of the torque, becomes unstable. These
calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal, fabrycky_otides}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta$ for
    the four CSs. The solid lines give $\lambda_0^2$ for the unperturbed
    $\phi_{\rm cs}$, and the dashed lines give the values for $\phi_{\rm cs}$
    shifted by $60^\circ$ ($\phi_{\rm cs} = 120^\circ$ for CS2 and $\phi_{\rm
    cs} = 60^\circ$ for CSs 1, 3, and 4).
    }\label{fig:lambda_full}
\end{figure}

@@

With the above results, we are equipped to ask questions about the dynamics of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}): what is the long term behavior for
a general initial $\uv{s}$? If $\eta > \eta_{\rm c}$, then the only possible
final spin state is CS2, and all initial conditions will evolve asympotically
towards it. As such, we consider only $\eta < \eta_{\rm c}$, where an initial
condition can asymptotically evolve towards either CS1 or CS2. We numerically
integrate Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) for many random initial
conditions uniformly distributed in $\p{\cos \theta, \phi}$ with $\abs{g}t_{\rm
s} = 10^{-3}$ and record the nearest CS for each integration after $10t_{\rm
s}$. In Fig.~\ref{fig:toy_phop}, we show the results of this procedure for $\eta
= 0.2$, and $I = 20^\circ$. It is clear that initial conditions in zone I evolve
into CS1, those in zone II evolve into CS2, while those in zone III have a
probabilistic outcome. We aim to understand each of these in turn:
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{Plot illustrating the asymptotic behavior of initial conditions for
    $\eta = 0.2$ and $I = 20^\circ$. Each dot represents an initial spin
    orientation, and the coloring of the dot indicates which Cassini State
    (legend) the system asymptotes towards.}\label{fig:toy_phop}
\end{figure}

@@

To understand this process quantitatively, as well as associate probabilities to
the two possible outcomes, it is important to be more quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves with the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx g\sin I + \frac{g^2}{2\alpha}\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we can compute $\Delta H_{\pm}$ the
change over both legs of the separatrix with
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm s}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
Thus, if we evaluate $H$ every time that a trajectory crosses $\phi = 0$, we see
that it will initially be negative and increase for each circulation cycle until
the system encounters the separatrix. At the beginning of the
separatrix-crossing orbit, the initial value of $H$, denoted $H_{\rm i}$,
satisfies $H_{\rm sep} - \Delta H_- \leq H_{\rm i} \leq H_{\rm sep}$: if it is
less than the lower bound, it will not cross the separatrix on this orbit, and
if it is greater than the upper bound then it is already inside the separatrix.
The two endpoints of this range are denoted in Fig.~\ref{fig:6equils} by the
black dot and cross at the left and right edges of the plot.

During the separatrix-crossing orbit, the trajectory first evolves
approximately along $\mathcal{C}_-$ then along $\mathcal{C}_+$, after which two
possible outcomes can occur:
\begin{itemize}
    \item If the final value of $H$ at the end of this separatrix traversal,
        denoted $H_{\rm f}$, satisfies $H_{\rm f} > H_{\rm sep}$, then the
        trajectory enters the separatrix, following the red shaded region in
        Fig.~\ref{fig:6equils}, and executes a zone III $\to$ II transition.
